An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-15 manuscript.Rmd
small edits

Existing proposals to address the challenges 1-3 either involve adopting
an relatively new workflow that may be foreign to both the researcher and
their collaborators, as in the case of many workflow software solutions,
or adopt a relatively black-box approach that cannot be easily extended
by other researchers, as in the case of virtual machines.

------------------------------------

<!--
# Docker

At the heart of the computational reproducibility challenge is the 
complexity, diversity, and rapidly changing nature of the computer software
that underpin such work.

# The alternatives

<!-- Current approaches at the Operating Systems level -->

@@

In short, workflow software expects a new approach to computational research

In contrast, virtual machines (VMs) offer a far more brute-force approach.  If the goal
is to reproduce everything as it ran on the researcher's own computer, well 
then we can just take a snapshot of the entire computer: data, software, 
operating system and all. To make this practical, [Dudley 2013] and [Howe 2012]
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

@@

## A "DevOpts" approach

The problems highlighted here are not unique to academic software, but
impact software development in general. While the academic research 
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual machines,
the software development community has recently emphasized a philosophy 
(rather than a particular tool), known as _Development_ and Systems _Operation_, or more 
frequently just "DevOpts." The approach is characterized by scripting,
rather than documenting, a description of the necessary dependencies 
for software to run, usually from the Operating System (OS) on up.

[Clark 2014] describes the DevOpts approach
along with both its relevance to reproducible research and examples of
its use in the academic research context.

Their paper provides both a consise statement of the problem we
have already discussed and the solution provided under this banner:

@@

The authors contrast this to the DevOpts approach:

@@

which elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. [Clark 2014] are careful to note
that this is not so much a technological shift as a philosophical one:

@@

I focus the remainder of this paper on one of the most recent and rapidly
growing among these, called Docker, which offers several promising
developments above and beyond the tools highlighted in [Clark 2014]
in providing a DevOpts approach to reproducible research. My goal
in focusing on this example is not to promote a particular solution, 
but to anchor the discussion of technical solutions to reproducibility
challenges in concrete examples.

@@

- Portable deployment: Docker defines a format for bundling and exchanging
containers such that they run identically across platforms.

- Automatic builds: Docker defines a format (Dockerfiles) that provide
a simple, scripted way to construct containers, following the DevOpts
philosophy.

- Versioning: Docker provides git-like capibilities for tracking changes
to a container, inspecting the history, metadata, or diff between
versions, committing versions, rolling back changes, and incremential
uploads and downloads that send only the differences (like `git push`
or `git pull`).

- Component reuse: Any container can be used as the base for any other.

- Sharing: Docker provides a public registry (Docker Hub) where any
user can make images avaialble to any other. (The registry itself is
also open source so that it can be deployed privately).

@@

@@

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.
A researcher more familiar with python could connect our MySQL container
to their python container, and so forth.

@@

@@

[Clark 2014]

>>>>

<!-- Obviously reproducible research barriers go much deeper than this,
code that is not modular, not documented, not functionalized or otherwise
easy to read and follow are all obstacles to resuse.-->

<!-- A wide range of proposals, we focus on those at the operating system level -->
Though a wide variety of approaches exists to work around these challenges,
few operate on a low enough level to provide a general solution.  [Clark 2014]
provide an excellent description of this situation:

There are two dominant approaches to this issue of coordinating multiple
tools: Workflows and Virtual Machines (VMs).

# Current approaches

@@

In short, workflow software expects a new approach to computational research. 
In contrast, virtual machines (VMs) offer a more direct approach. Since
the computer Operating System (OS) already provides the software layer
responsible for coodinating all the different elements running on the computer,
the VM approach captures the OS and everything running on it whole-cloth.
To make this practical, [Dudley 2013] and [Howe 2012]
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

@@

<!-- Scripted provisioning: a solution to the weaknesses of virtual machines -->
## A "DevOpts" approach

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOpts." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up.  [Clark 2014] describes the DevOpts approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
we have discussed so far in terms of effective documentation:

@@

The authors contrast this to the DevOpts approach, where dependency 
documentation is _scripted_:

@@

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. [Clark 2014] are careful to note
that this is not so much a technological shift as a philosophical one:

@@

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in [Clark
2014]. Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

@@

- *Portable deployment*: Docker defines a format for bundling and
exchanging containers such that they run identically across platforms.

- *Automatic builds*: Docker defines a format (Dockerfiles) that provide
a simple, scripted way to construct containers, following the DevOpts
philosophy.

- *Versioning*: Docker provides git-like capibilities for tracking
changes to a container, inspecting the history, metadata, or diff between
versions, committing versions, rolling back changes, and incremential
uploads and downloads that send only the differences (like `git push`
or `git pull`).

- *Component reuse*: Any container can be used as the base for any other.

- *Sharing*: Docker provides a public registry (Docker Hub) where any
user can make images avaialble to any other. (The registry itself is
also open source so that it can be deployed privately).

@@

1. **"Dependency Hell"**
2.  **Imprecise documentation**
3.  **Code rot**
4. **Barriers to adoption and reuse in existing solutions**

------

@@

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.
A researcher could connect a container providing a computational 
environment for a different language to this same MySQL container, 
and so forth.

@@

<!-- excerpted -->

@@
